\texttt{MSnbase}, efficient and elegant R-based processing and visualisation of raw mass spectrometry data

### 2020-07-15 msnbase2.Rnw
merge PRs, compile with red text

\documentclass[journal=jacsat,manuscript=article]{achemso}\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

@@

Practitioners have a diverse catalogue of software tools to explore,
process and interpret mass spectrometry data at their disposal. These
range from low level software libraries that are aimed at programmers
to enable development of new applications, to more user-oriented
applications with graphical user interfaces which provide a more
limited set of functionalities to address a defined scope. Examples of
software libraries include Java-based jmzML~\cite{Cote:2010} or
C/C++-based ProteoWizard~\cite{Chambers:2012}.
\sout{ProteomeDiscoverer (Thermo Scientific)}
\red{Thermo Scientific Proteome Discoverer (Thermo Fisher Scientific)},
MaxQuant~\cite{Cox:2008} and PeptideShaker~\cite{Vaudel:2015} are among the
most widely used user-centric applications.

In this software note, we present version 2 of the
\texttt{MSnbase}~\cite{Gatto:2012} software, available from the
Bioconductor~\cite{Huber:2015} project. The package, like other
software such as Python-based {pyOpenMS}~\cite{Rost:2014},
spectrum\_utils~\cite{Bittremieux:2020} or
Pyteomics~\cite{Goloborodko:2013}, offers a platform that lies between
low level libraries and end-user software. \texttt{MSnbase} provides a
flexible R~\cite{R} command-line environment for metabolomics and
proteomics mass spectrometry-based applications. It lays out a sound
infrastructure to work with raw mass spectrometry \color{red}{data from MS files
  in mzML, mzXML, mzData or ANDI-MS/netCDF format as well as}\sout{,} quantitative and
proteomics identification data. The package enables manipulation (for
example subsetting, filtering, or accessing specific parts thereof),
detailed step-by-step processing (for example smoothing and
centroiding of \sout{raw} \color{red}{profile-mode MS} data, or normalisation and imputation of
quantitative data), analysis and visualisation of these data and the
development of novel computational mass spectrometry
methods~\cite{Stanstrup:2019}. Here, we focus on the new developments
pertaining to raw mass spectrometry data handling and processing.

@@

In \texttt{MSnbase}, mass spectrometry experiments are handled as
\texttt{MSnExp} objects. While the implementation is more complex, it
is useful to schematise a raw data experiment as being composed of raw
data, i.e. a collection of individual spectra, as well as
spectra-level metadata. Each spectrum is composed of m/z values and
associated intensities. The metadata are represented by a table with
variables along the columns and each row associated to a
spectrum. Among the metadata available for each spectrum, there are MS
level, acquisition number, retention time, precursor m/z and intensity
(for MS level 2 and above), and many more. \texttt{MSnbase}\color{red}{
  ,\textit{via} the \texttt{mzR} package~\cite{Chambers:2012}, allows import of raw mass spectrometry
  data from files in mzML, mzXML, mzdata or ANDI-MS/netCDF format and} provides a
rich interface to manipulate such objects. The code chunk below
illustrates such an object as displayed in the R console and an
enumeration of the metadata fields.

@@

In the following sections, we describe \sout{\texttt{MSnbase}'s ability to
  efficiently handle very large mass spectrometry data and experiments and how
  it} \code{red}{how \texttt{MSnbase}} can be used for data processing and
visualisation. We will also illustrate how it makes use of the forward-pipe
operator (\texttt{\%>\%}) defined in the \texttt{magrittr} package. This
operator has proved useful to develop non-trivial analyses by combining
individual functions into easily readable pipelines. \code{red}{An example for
  \texttt{MSnbase}'s ability to efficiently handle also very large mass
  spectrometry experiments (in this case with 5,773,464 spectra in 1,182 mzXML
  files) is provided as Supplementary Information.}

@@

The main feature in version 2 of the \texttt{MSnbase} package was the
addition of different backends for raw data storage, namely
\textit{in-memory} and \textit{on-disk}. The following code chunk
demonstrates how to create two \texttt{MSnExp} objects \color{red}{importing
  data from an mzML file and} storing \sout{data}\color{red}{it} 
either in-memory or on-disk.

@@

\color{red}{Because the on-disk backend does not hold all the data in memory,
  direct data manipulations are not possible. We thus implemented a \textit{lazy
  processing} mechanism in this backend that caches any data manipulation
operation in a processing queue in the object itself. Only when the user
accesses m/z or intensity values these operations are applied}.
\sout{The on-disk backend also offers efficient data manipulation by way of
\textit{lazy processing} in the object itself. Operations on the raw
data are stored in a processing queue and only effectively applied
when the user accesses m/z or intensity values. \color{red}{As an additional
  advantage, operations on subsets of the data become much faster since data
  manipulations are applied only to data subsets instead of the full data set
  at once. Also, data access in on-disk backends is parallelized by data file
  ensuring a higher performance of this backend over conventional in-memory data
  representations.} 
As an example, the
following short analysis pipeline, that can equally be applied to
in-memory or on-disk data, retains MS2 spectra acquired between 1000
and 3000 seconds, extracts the m/z range corresponding to the TMT
6-plex range and focuses on the MS2 spectra with a precursor intensity
greater than $11 \times 10^6$ (the median precursor intensity).

@@

The R environment is well known for the quality of its visualisation
capacity. This also holds true for mass
spectrometry\cite{Gatto:2015,protViz,Pviz,Bemis:2015}. Here, we
conclude the overview of version 2 of the \texttt{MSnbase} package by
highlighting the flexibility of the software to visualise and assess
the efficiency of raw data processing. Figure \ref{fig:cent} compares
the raw MS profile data \color{red}{imported from an mzML file} for \sout{the} serine and the same data after
smoothing, centroiding and m/z refinement, as illustrated in the code
chunk below. Detailed execution and description of these operations
can be found in the \emph{MSnbase: centroiding of profile-mode MS
  data} \texttt{MSnbase} vignette.

>>>>

\documentclass[journal=jacsat,manuscript=article]{achemso}

\usepackage[]{graphicx}
\usepackage[]{color}

\usepackage[]{xcolor}
\usepackage[normalem]{ulem}

%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

@@

Practitioners have a diverse catalogue of software tools to explore,
process and interpret mass spectrometry data at their disposal. These
range from low level software libraries that are aimed at programmers
to enable development of new applications, to more user-oriented
applications with graphical user interfaces which provide a more
limited set of functionalities to address a defined scope. Examples of
software libraries include Java-based jmzML~\cite{Cote:2010} or
C/C++-based ProteoWizard~\cite{Chambers:2012}.
\sout{ProteomeDiscoverer (Thermo Scientific)}
\textcolor{red}{Thermo Scientific Proteome Discoverer (Thermo Fisher Scientific)},
MaxQuant~\cite{Cox:2008} and PeptideShaker~\cite{Vaudel:2015} are among the
most widely used user-centric applications.

In this software note, we present version 2 of the
\texttt{MSnbase}~\cite{Gatto:2012} software, available from the
Bioconductor~\cite{Huber:2015} project. The package, like other
software such as Python-based {pyOpenMS}~\cite{Rost:2014},
spectrum\_utils~\cite{Bittremieux:2020} or
Pyteomics~\cite{Goloborodko:2013}, offers a platform that lies between
low level libraries and end-user software. \texttt{MSnbase} provides a
flexible R~\cite{R} command-line environment for metabolomics and
proteomics mass spectrometry-based applications. It lays out a sound
infrastructure to work with raw mass spectrometry \textcolor{red}{data from MS files
  in mzML, mzXML, mzData or ANDI-MS/netCDF format as well as}\sout{,} quantitative and
proteomics identification data. The package enables manipulation (for
example subsetting, filtering, or accessing specific parts thereof),
detailed step-by-step processing (for example smoothing and
centroiding of \sout{raw} \textcolor{red}{profile-mode MS} data, or normalisation and imputation of
quantitative data), analysis and visualisation of these data and the
development of novel computational mass spectrometry
methods~\cite{Stanstrup:2019}. \textcolor{red}{Extensive documentation and use cases 
are provided in \textit{package vignettes}~\cite{MSnbaseVignettes} and 
workflows~\cite{xcmsWorkflow}.} Here, we focus on the new developments
pertaining to raw mass spectrometry data handling and processing.

@@

In \texttt{MSnbase}, mass spectrometry experiments are handled as
\texttt{MSnExp} objects. While the implementation is more complex, it
is useful to schematise a raw data experiment as being composed of raw
data, i.e. a collection of individual spectra, as well as
spectra-level metadata. Each spectrum is composed of m/z values and
associated intensities. The metadata are represented by a table with
variables along the columns and each row associated to a
spectrum. Among the metadata available for each spectrum, there are MS
level, acquisition number, retention time, precursor m/z and intensity
(for MS level 2 and above), and many
more. \texttt{MSnbase}\textcolor{red}{, \textit{via} the \texttt{mzR}
  package~\cite{Chambers:2012}, allows import of raw mass spectrometry
  data from files in mzML, mzXML, mzdata or ANDI-MS/netCDF format and}
provides a rich interface to manipulate such objects. The code chunk
below illustrates such an object as displayed in the R console and an
enumeration of the metadata fields.

@@

In the following sections, we describe \sout{\texttt{MSnbase}'s ability to
  efficiently handle very large mass spectrometry data and experiments and how
  it} \textcolor{red}{how \texttt{MSnbase}} can be used for data processing and
visualisation. We will also illustrate how it makes use of the forward-pipe
operator (\texttt{\%>\%}) defined in the \texttt{magrittr} package. This
operator has proved useful to develop non-trivial analyses by combining
individual functions into easily readable pipelines. \textcolor{red}{An example for
  \texttt{MSnbase}'s ability to efficiently handle also very large mass
  spectrometry experiments (in this case with 5,773,464 spectra in 1,182 mzXML
  files) is provided as Supplementary Information.}

@@

The main feature in version 2 of the \texttt{MSnbase} package was the
addition of different backends for raw data storage, namely
\textit{in-memory} and \textit{on-disk}. The following code chunk
demonstrates how to create two \texttt{MSnExp} objects \textcolor{red}{importing
  data from an mzML file and} storing \sout{data}\textcolor{red}{it} 
either in-memory or on-disk.

@@

\textcolor{red}{Because the on-disk backend does not hold all the data in memory,
  direct data manipulations are not possible. We thus implemented a \textit{lazy
  processing} mechanism in this backend that caches any data manipulation
operation in a processing queue in the object itself. Only when the user
accesses m/z or intensity values these operations are applied}.
\sout{The on-disk backend also offers efficient data manipulation by way of
\textit{lazy processing} in the object itself.} Operations on the raw
data are stored in a processing queue and only effectively applied
when the user accesses m/z or intensity values. \textcolor{red}{As an additional
  advantage, operations on subsets of the data become much faster since data
  manipulations are applied only to data subsets instead of the full data set
  at once. Also, data access in on-disk backends is parallelized by data file
  ensuring a higher performance of this backend over conventional in-memory data
  representations.} 
As an example, the
following short analysis pipeline, that can equally be applied to
in-memory or on-disk data, retains MS2 spectra acquired between 1000
and 3000 seconds, extracts the m/z range corresponding to the TMT
6-plex range and focuses on the MS2 spectra with a precursor intensity
greater than $11 \times 10^6$ (the median precursor intensity).

@@

The R environment is well known for the quality of its visualisation
capacity. This also holds true for mass
spectrometry\cite{Gatto:2015,protViz,Pviz,Bemis:2015}. Here, we
conclude the overview of version 2 of the \texttt{MSnbase} package by
highlighting the flexibility of the software to visualise and assess
the efficiency of raw data processing. Figure \ref{fig:cent} compares
the raw MS profile data \textcolor{red}{imported from an mzML file} for \sout{the} serine and the same data after
smoothing, centroiding and m/z refinement, as illustrated in the code
chunk below. Detailed execution and description of these operations
can be found in the \emph{MSnbase: centroiding of profile-mode MS
  data} \texttt{MSnbase} vignette.

### 2020-07-15 msnbase2.tex
merge PRs, compile with red text

\usepackage{alltt}\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

@@

Practitioners have a diverse catalogue of software tools to explore,
process and interpret mass spectrometry data at their disposal. These
range from low level software libraries that are aimed at programmers
to enable development of new applications, to more user-oriented
applications with graphical user interfaces which provide a more
limited set of functionalities to address a defined scope. Examples of
software libraries include Java-based jmzML~\cite{Cote:2010} or
C/C++-based ProteoWizard~\cite{Chambers:2012}. ProteomeDiscoverer
(Thermo Scientific), MaxQuant~\cite{Cox:2008} and
PeptideShaker~\cite{Vaudel:2015} are among the most widely used
user-centric applications.

In this software note, we present version 2 of the
\texttt{MSnbase}~\cite{Gatto:2012} software, available from the
Bioconductor~\cite{Huber:2015} project. The package, like other
software such as Python-based {pyOpenMS}~\cite{Rost:2014},
spectrum\_utils~\cite{Bittremieux:2020} or
Pyteomics~\cite{Goloborodko:2013}, offers a platform that lies between
low level libraries and end-user software. \texttt{MSnbase} provides a
flexible R~\cite{R} command-line environment for metabolomics and
proteomics mass spectrometry-based applications. It lays out a sound
infrastructure to work with raw mass spectrometry, quantitative and
proteomics identification data. The package enables manipulation (for
example subsetting, filtering, or accessing specific parts thereof),
detailed step-by-step processing (for example smoothing and
centroiding of raw data, or normalisation and imputation of
quantitative data), analysis and visualisation of these data and the
development of novel computational mass spectrometry
methods~\cite{Stanstrup:2019}. Here, we focus on the new developments
pertaining to raw mass spectrometry data handling and processing.

@@

In \texttt{MSnbase}, mass spectrometry experiments are handled as
\texttt{MSnExp} objects. While the implementation is more complex, it
is useful to schematise a raw data experiment as being composed of raw
data, i.e. a collection of individual spectra, as well as
spectra-level metadata. Each spectrum is composed of m/z values and
associated intensities. The metadata are represented by a table with
variables along the columns and each row associated to a
spectrum. Among the metadata available for each spectrum, there are MS
level, acquisition number, retention time, precursor m/z and intensity
(for MS level 2 and above), and many more. \texttt{MSnbase} provides a
rich interface to manipulate such objects. The code chunk below
illustrates such an object as displayed in the R console and an
enumeration of the metadata fields.

@@

In the following sections, we describe \texttt{MSnbase}'s ability to
efficiently handle very large mass spectrometry data and experiments
and how it can be used for data processing and visualisation. We will
also illustrate how it makes use of the forward-pipe operator
(\texttt{\%>\%}) defined in the \texttt{magrittr} package. This
operator has proved useful to develop non-trivial analyses by
combining individual functions into easily readable pipelines.

@@

The main feature in version 2 of the \texttt{MSnbase} package was the
addition of different backends for raw data storage, namely
\textit{in-memory} and \textit{on-disk}. The following code chunk
demonstrates how to create two \texttt{MSnExp} objects storing data
either in-memory or on-disk.

@@

The on-disk backend also offers efficient data manipulation by way of
\textit{lazy processing} in the object itself. Operations on the raw
data are stored in a processing queue and only effectively applied
when the user accesses m/z or intensity values. As an example, the
following short analysis pipeline, that can equally be applied to
in-memory or on-disk data, retains MS2 spectra acquired between 1000
and 3000 seconds, extracts the m/z range corresponding to the TMT
6-plex range and focuses on the MS2 spectra with a precursor intensity
greater than $11 \times 10^6$ (the median precursor intensity).

@@

The R environment is well known for the quality of its visualisation
capacity. This also holds true for mass
spectrometry\cite{Gatto:2015,protViz,Pviz,Bemis:2015}. Here, we
conclude the overview of version 2 of the \texttt{MSnbase} package by
highlighting the flexibility of the software to visualise and assess
the efficiency of raw data processing. Figure \ref{fig:cent} compares
the raw MS profile data for the serine and the same data after
smoothing, centroiding and m/z refinement, as illustrated in the code
chunk below. Detailed execution and description of these operations
can be found in the \emph{MSnbase: centroiding of profile-mode MS
  data} \texttt{MSnbase} vignette.

@@

The version of \texttt{MSnbase} used in this manuscript is
2.14.0. The main features presented here were available since
version 2.0. The code to reproduce the analyses and figures in this
article is available at
\url{https://github.com/lgatto/2020-msnbase-v2/}.

>>>>

\usepackage{alltt}

\usepackage[]{graphicx}
\usepackage[]{color}

\usepackage[]{xcolor}
\usepackage[normalem]{ulem}

%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

@@

Practitioners have a diverse catalogue of software tools to explore,
process and interpret mass spectrometry data at their disposal. These
range from low level software libraries that are aimed at programmers
to enable development of new applications, to more user-oriented
applications with graphical user interfaces which provide a more
limited set of functionalities to address a defined scope. Examples of
software libraries include Java-based jmzML~\cite{Cote:2010} or
C/C++-based ProteoWizard~\cite{Chambers:2012}.
\sout{ProteomeDiscoverer (Thermo Scientific)}
\textcolor{red}{Thermo Scientific Proteome Discoverer (Thermo Fisher Scientific)},
MaxQuant~\cite{Cox:2008} and PeptideShaker~\cite{Vaudel:2015} are among the
most widely used user-centric applications.

In this software note, we present version 2 of the
\texttt{MSnbase}~\cite{Gatto:2012} software, available from the
Bioconductor~\cite{Huber:2015} project. The package, like other
software such as Python-based {pyOpenMS}~\cite{Rost:2014},
spectrum\_utils~\cite{Bittremieux:2020} or
Pyteomics~\cite{Goloborodko:2013}, offers a platform that lies between
low level libraries and end-user software. \texttt{MSnbase} provides a
flexible R~\cite{R} command-line environment for metabolomics and
proteomics mass spectrometry-based applications. It lays out a sound
infrastructure to work with raw mass spectrometry \textcolor{red}{data from MS files
  in mzML, mzXML, mzData or ANDI-MS/netCDF format as well as}\sout{,} quantitative and
proteomics identification data. The package enables manipulation (for
example subsetting, filtering, or accessing specific parts thereof),
detailed step-by-step processing (for example smoothing and
centroiding of \sout{raw} \textcolor{red}{profile-mode MS} data, or normalisation and imputation of
quantitative data), analysis and visualisation of these data and the
development of novel computational mass spectrometry
methods~\cite{Stanstrup:2019}. \textcolor{red}{Extensive documentation and use cases 
are provided in \textit{package vignettes}~\cite{MSnbaseVignettes} and 
workflows~\cite{xcmsWorkflow}.} Here, we focus on the new developments
pertaining to raw mass spectrometry data handling and processing.

@@

In \texttt{MSnbase}, mass spectrometry experiments are handled as
\texttt{MSnExp} objects. While the implementation is more complex, it
is useful to schematise a raw data experiment as being composed of raw
data, i.e. a collection of individual spectra, as well as
spectra-level metadata. Each spectrum is composed of m/z values and
associated intensities. The metadata are represented by a table with
variables along the columns and each row associated to a
spectrum. Among the metadata available for each spectrum, there are MS
level, acquisition number, retention time, precursor m/z and intensity
(for MS level 2 and above), and many
more. \texttt{MSnbase}\textcolor{red}{, \textit{via} the \texttt{mzR}
  package~\cite{Chambers:2012}, allows import of raw mass spectrometry
  data from files in mzML, mzXML, mzdata or ANDI-MS/netCDF format and}
provides a rich interface to manipulate such objects. The code chunk
below illustrates such an object as displayed in the R console and an
enumeration of the metadata fields.

@@

In the following sections, we describe \sout{\texttt{MSnbase}'s ability to
  efficiently handle very large mass spectrometry data and experiments and how
  it} \textcolor{red}{how \texttt{MSnbase}} can be used for data processing and
visualisation. We will also illustrate how it makes use of the forward-pipe
operator (\texttt{\%>\%}) defined in the \texttt{magrittr} package. This
operator has proved useful to develop non-trivial analyses by combining
individual functions into easily readable pipelines. \textcolor{red}{An example for
  \texttt{MSnbase}'s ability to efficiently handle also very large mass
  spectrometry experiments (in this case with 5,773,464 spectra in 1,182 mzXML
  files) is provided as Supplementary Information.}

@@

The main feature in version 2 of the \texttt{MSnbase} package was the
addition of different backends for raw data storage, namely
\textit{in-memory} and \textit{on-disk}. The following code chunk
demonstrates how to create two \texttt{MSnExp} objects \textcolor{red}{importing
  data from an mzML file and} storing \sout{data}\textcolor{red}{it} 
either in-memory or on-disk.

@@

\textcolor{red}{Because the on-disk backend does not hold all the data in memory,
  direct data manipulations are not possible. We thus implemented a \textit{lazy
  processing} mechanism in this backend that caches any data manipulation
operation in a processing queue in the object itself. Only when the user
accesses m/z or intensity values these operations are applied}.
\sout{The on-disk backend also offers efficient data manipulation by way of
\textit{lazy processing} in the object itself.} Operations on the raw
data are stored in a processing queue and only effectively applied
when the user accesses m/z or intensity values. \textcolor{red}{As an additional
  advantage, operations on subsets of the data become much faster since data
  manipulations are applied only to data subsets instead of the full data set
  at once. Also, data access in on-disk backends is parallelized by data file
  ensuring a higher performance of this backend over conventional in-memory data
  representations.} 
As an example, the
following short analysis pipeline, that can equally be applied to
in-memory or on-disk data, retains MS2 spectra acquired between 1000
and 3000 seconds, extracts the m/z range corresponding to the TMT
6-plex range and focuses on the MS2 spectra with a precursor intensity
greater than $11 \times 10^6$ (the median precursor intensity).

@@

The R environment is well known for the quality of its visualisation
capacity. This also holds true for mass
spectrometry\cite{Gatto:2015,protViz,Pviz,Bemis:2015}. Here, we
conclude the overview of version 2 of the \texttt{MSnbase} package by
highlighting the flexibility of the software to visualise and assess
the efficiency of raw data processing. Figure \ref{fig:cent} compares
the raw MS profile data \textcolor{red}{imported from an mzML file} for \sout{the} serine and the same data after
smoothing, centroiding and m/z refinement, as illustrated in the code
chunk below. Detailed execution and description of these operations
can be found in the \emph{MSnbase: centroiding of profile-mode MS
  data} \texttt{MSnbase} vignette.

@@

The version of \texttt{MSnbase} used in this manuscript is
2.14.2. The main features presented here were available since
version 2.0. The code to reproduce the analyses and figures in this
article is available at
\url{https://github.com/lgatto/2020-msnbase-v2/}.
